Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-05-12 Manuscript/PLoSONEresponse/TEM_PLoS_revision.tex
Updated final response

\subsection*{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \cite{R302} using the package diversitree v. 0.9-6 \cite{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \cite{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \cite{beckancient2014}, 58\% \cite{schragocombining2013}, 48\% \cite{pyrondivergence2011}, 31\% \cite{ronquista2012} and 31\% \cite{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. The fossil taxa were considered as unique tips at the end of extinct lineages. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data. Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa. We first applied the parameters $M_L$ and $M_F$ on the matrix and then applied the $M_C$ parameter. Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix. However, when applying the $M_C$ parameter with the same rate (10\%) the resulting matrix potentially had more than 10\% data missing.

@@

\paragraph*{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \cite{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \cite{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \cite{Stamatakis21012014}). For the morphological data, we used the implemented M\textit{kv} model \cite{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \cite{Stamatakis21012014}). We used RAxML because it automatically corrects for acquisition bias \cite{lewisa2001}. It also is heavily used in the literature for Maximum Likelihood tree inference (e.g. \cite{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \cite{Stamatakis01102008}.

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \cite{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). This step was implemented to make sure that the differences we observed in topologies (see below) where due to the amount of missing data for each parameters ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies. On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \cite{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric following Bogdanowicz's Normalised Tree Similarity (NTS) method \cite{Bogdanowicz2012}. This methods scales any tree comparison metric using the mean distance between 1000 random trees (see ~\nameref{S2_Appendix2} for the calculation details). This method is a generalisation of the topological accuracy method \cite{Price2010} allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see ~\nameref{S2_Appendix2} for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the ``best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule. When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxon in the clade that shares most characters with the fossils. Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade. 
In this case, the Normalised Robinson-Foulds metric will decreases when the fossil are present at a lower taxonomic level (i.e. separated with more nodes from the root) but affects the clade conservation less at higher taxonomic level (i.e. separated with less nodes from the root). On the opposite though, because the fossil in a high level clade has many chances to branch on different nodes within that clade, it will be more likely to act as a wildcard taxa and decrease the Normalised Triplets metric. Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees.
%This does not decrease the Normalised Robinson-Foulds metric, because clades are conserved, but decreases the Normalised Triplets metric because the fossils act as wildcard taxa.
Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50). In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level. We argue that using the Bayesian consensus tree topology is preferable because it is more conservative (e.g. \cite{pattinsonphylogeny2014}).

>>>>

\subsection*{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \cite{R302} using the package diversitree v. 0.9-6 \cite{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution (bounded between 0 and 1) but maintaining $\lambda$ $>$ $\mu$ \cite{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \cite{beckancient2014}, 58\% \cite{schragocombining2013}, 48\% \cite{pyrondivergence2011}, 31\% \cite{ronquista2012} and 31\% \cite{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. The fossil taxa were considered as unique tips at the end of extinct lineages. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data. Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa. We first applied the parameters $M_L$ and $M_F$ to the matrix and then applied the $M_C$ parameter. Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix. However, when applying the $M_C$ parameter with the same percentage (10\%) the resulting matrix potentially had more than 10\% data missing.

@@

\paragraph*{Maximum Likelihood}
The ``best" tree and the ``missing-data" trees were inferred using RAxML v. 8.0.20 \cite{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model \cite{tavare1986} (default GTRGAMMA in RAxML v. 8.0.20  \cite{Stamatakis21012014}). For the morphological data, we used the implemented M\textit{kv} model \cite{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20 \cite{Stamatakis21012014}). We used RAxML because it automatically corrects for acquisition bias \cite{lewisa2001}. It is also heavily used in the literature for Maximum Likelihood tree inference (e.g. \cite{rouresite-specific2011,Bogdanowicz2012,springermacroevolutionary2012,O'Leary08022013,kellymolecular2014}) and is one of the fastest methods available \cite{Stamatakis01102008}.

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \cite{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete" and 50 x 125 ``missing-data" matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50). This step was implemented to make sure that the differences we observed in topologies (see below) were due to the amount of missing data for each parameter ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies. On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \cite{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric following Bogdanowicz's Normalised Tree Similarity (NTS) method \cite{Bogdanowicz2012}. This methods scales any tree comparison metric using the mean distance between 1000 random trees (see ~\nameref{S2_Appendix2} for the calculation details). This method is a generalisation of the topological accuracy method \cite{Price2010} allowing to compare topological differences between any tree with any tree comparison metric.In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance. Note that once rescaled, the Normalised Robinson-Foulds metric is a measure of similarity, rather than of distance like the original Robinson-Foulds metric.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see ~\nameref{S2_Appendix2} for calculation details). Note that this is comparable to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the ``best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

The effect of the missing data in the fossil record ($M_{F}$) is less than the effect of the $M_{L}$ parameter on clade conservation (Normalised Robinson-Foulds metric) but greater on the displacement of wildcard taxa (Normalised Triplets metric; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). This is related to the fact that the Bayesian consensus tree is built using a majority consensus rule. When the fossil taxa have less data (e.g. $M_{F}$ = 75\%) they will tend to branch with any taxon in the clade that shares most characters with the fossils. Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees in the Bayesian posterior distribution) and the fossil taxa will form a polytomy at the base of the clade. In this case, the Normalised Robinson-Foulds metric will decrease when the fossil is present at a lower taxonomic level (i.e. separated with more nodes from the root) but affects the clade conservation less at higher taxonomic level (i.e. separated with less nodes from the root). Conversely, because a fossil in a high taxonomic level clade has many chances to branch on different nodes within the clade, it will be more likely to act as a wildcard taxa and decrease the Normalised Triplets metric. Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees. Conversely, the same scenario in a Maximum Likelihood framework will lead to a dichotomous branching of the fossils but with low bootstrap support ($<$ 50). In other words, the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree, where the fossil will be placed with lower confidence at a higher taxonomic level. We argue that using the Bayesian consensus tree topology is preferable because it is more conservative (e.g. \cite{pattinsonphylogeny2014}).
